Notes and other text

### 2023-04-14 doc/manuscript/translocation.qmd
Edits to first paragraph of Discussion, other minor changes

To detect regions of the genome under selection, we used three complementary approaches that together describe the evolutionary response of frog populations to Bd.
First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and individual variants.
For these single nucleotide polymorphisms (SNPs), we identified outlier variants using two Bonferroni-corrected significance levels: "liberal" = 0.05, "stringent" = 0.01.
Second, we used a splined window approach to evaluate larger genomic regions showing differentiation between population types in *F~ST~* and nucleotide diversity ($\pi$) <!--# Allie comment: Feels like we need to signal here that we are calculating the difference in pi between naive and persistent. pi-diff.  -->.
In this analysis, we identified outlier regions using a threshold for the W-statistic (4 standard deviations from the mean) and selected regions where outlier windows overalapped for *F~ST~* and $\pi$ <!--# This is a different definition than the one given in the Methods: "4 * standard deviation". --> <!--# Allie comment: Feels like we need to signal here that we are calculating the difference in pi between naive and persistent. pi - diff -->.
Third, we assessed whether putative genes or genomic regions that showed strong evidence of selection were associated with biological functions, molecular functions, or cellular processes (i.e., GO terms) consistent with possible resistance mechanisms.

@@

Our second analysis, using a splined window approach, identified regions of the genome that had both high differentiation (*F~ST~*) and a large difference in nucleotide diversity between naive and recovering populations ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).
We identified 58 outlier windows for *F~ST~* and 33 outlier windows for $\pi_{diff}$ (@fig-spline-manhattan A, B).
Of these, 9 regions were outliers for both metrics and two of these shared regions also contained outlier SNPs.<!--# "windows" and "regions" are used interchangeably in this paragraph. To avoid confusion, use "windows" throughout? --> A total of 33 annotated genes were found in the 9 shared regions (SI Dataset 2 <!--# In the supplementary_table_outliers_typofix.xlsx file > the splined_window_shared_ouliers sheet contains descriptions of 10 shared regions, not 9--> <!--# Allie comment: One is a repeat because there is a 2>1 match for one of the overlaps (i.e., One Fst outlier window overlaps with 2 pi outlier windows).  -->) and no GO terms were overrepresented within this set of genes.
The strongest signal for a reduction in $\pi$ in recovering populations (highest $\pi_{diff}$) <!--# Allie comment: For me this is enough of a reason to discuss this beyond the obvious relevance of the gene function.  -->, indicative of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
This region contains \~500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1 <!--# Allie comment: This gene is even more interesting now that I've done a lit review. More in the discussion! -->).
Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
Interestingly, this region had higher $\pi$ in the recovering populations, consistent with a pattern of balancing selection.
Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) <!--# Allie comment: Again, this is a very good reason to discuss - there are both outlier windows and individuals SNPS here making it an important part outlier region to discuss BEYOND the cool MHC connection.  -->and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D).
This region (854kb from the beginning of the outlier windows to the VARS SNP) contained a total of 10 annotated genes.
In *Xenopus*, four of these genes occur in the extended MHC Class I region (FLOT1, TUBB, MDC1,TCF19) and three occur in the extended MHC Class II region (HSP70, LSM2, VARS) [@ohta2006].
Therefore, we conclude that this candidate region under selection is a part of the extended MHC Class I and II complex and shows synteny with other amphibian genomes.
<!--# Here or in the discussion we will need to place this result in the context of previous work on MHC in amphibians.  Readers not familiar with amphibian-Bd systems might miss the importance of this result --><!--# Based on Fu and Waldman (2017), are any of our data sufficient to conclude anything related to statement that "Bd-resistant species share similar pocket conformations within the MHC-II antigen-binding groove. Among susceptible species, survivors of epizootics bear alleles encoding these conformations."? What about MHC Class II characteristics linked to resistance by Bataille et al. (2015)? Savage and Zamudio (2011, 2016), Kosch et al. (2018)?-->

@@

<!--# Need summary of importance of genetic results, also need better integration of different selection results.  Mark W: I took a crack at this to test my naive understanding of the genetic results. Figured I could be a good test case of how a genetically naive reader would summarize the key results, to see if our genomic message is coming across clearly. -->

@@

<!--# In reading through these results, I am not seeing a compelling theme that emerges from all of the different results, both within and between the SNP and splined window sections. The results show differences in various metrics between population types, but are these changes large or small? I lack the context to understand that. Similarly, is the magnitude of differences consistent with strong or weak selection? -->

@@

Disease-induced declines are a central conservation challenge for wildlife, and identifying strategies to recover affected species is a critical need.
One potential strategy, and the one we use in the current study, leverages disease-induced genetic changes in host populations and evolutionary rescue to recover species at the landscape scale.
Although previously proposed [@mendelson2019], to our knowledge this strategy has never been successfully implemented in wildlife disease systems [but see @joseph2018 for a small-scale example].
Here, we provide strong evidence that natural selection for resistance and the associated evolutionary rescue has facilitated recovery of populations of the endangered MYL frog in the presence of a deadly fungal pathogen.
The use of individuals from these resistant, rescued populations in a 15 year translocation effort plus associated population viability modeling demonstrate that populations that have experienced evolutionary rescue can be used to successfully establish and recover this species across the landscape.
These results not only provide a rare example of species recovery following disease-induced declines due heritable changes in resistance, but provide the first demonstration of how evolutionary rescue can be used to successfully recover a species in the presence of disease.

We provide three lines of evidence that support evolutionary rescue as a primary mechanism of recovery in MYL frog systems.
First, our genomic analyses comparing naive and recovering MYL frog populations found notable differences in allele frequencies between adjacent Bd-naive and recovering <!--# Check on using "persistent" vs "recovering" vs "recovered" --> <!--# Roland response: Use "recovering" throughout -->populations, consistent with strong selection.
Importantly, some of the regions of the genome that showed signatures of selection are associated with known cellular and immunological mechanisms contributing to disease resistance in other species generally [@robledo2018; @robledo2020] and amphibians specifically [@rosenblum2008; @ellison2014] (for more information, see SI X).
For example, we identified a \~940kb region on contig 8 that shows evidence of selection in recovering populations and contains eight genes associated with either the MHC Class I or Class III regions.
These results corroborate numerous previous studies linking MHC genes to Bd survival [e.g., @savage2011; @bataille2015].
Broadly, our genomic analyses imply that selection and rapid evolutionary change has contributed to survival in MYL frogs in the presence of Bd.
<!--# For brevity in the discussion, I tried to a pull a few key bullets out of Allie's text. Allie and Bree: Edit and expand as you see fit. -->

The second line of evidence that evolutionary rescue is a strong candidate mechanism for recovery is that repeated translocations and long-term monitoring of naturally recovered populations have demonstrated that resistant phenotypes are maintained beyond the natal environment in the presence of Bd (as reflected by the high year-one adult survival in 6 out of 12 translocated populations).
Moreover, resistance has a quantifiable phenotype in this system -- resistant, persisting populations have Bd infection intensities below the level known to cause significant host mortality [@vredenburg2010, @knapp2016].
Translocated populations maintain infection intensities below this threshold of infection intensity such that infection intensity is not a significant predictor of adult survival following translocations to new environments.
Third, high adult survival in the presence of Bd persists into subsequent generations in non-natal environments, suggesting that resistance is heritable to some degree.
Although previous studies have documented signatures of selection in hosts following disease-induced declines [@epstein2016, @gignoux-wolfsohn2021], our study is unique in that these three lines of evidence provide a strong case that evolved resistance begets population recovery in both natal and non-natal habitats, consistent with evolutionary rescue.

@@

1\) STRENGTH OF SIGNAL: the gene regions we discuss specifically in our dataset are robust outliers.
In the case of the MHC region we have multiple SNPs in this region that are statistically significantly associated AND a larger pattern of depressed pi in the recovering pops and large Fst.
In the case of GVINP1 we have the only instance of two adjacent outlier spline windows for pi, and the largest difference in pi of any region we measured.
Finally for RIN3 the allele frequency maps tell the most compelling story and reflect the fact that this is the most statistically significantly outlier SNP that has an annotation.
and

@@

>>>>

To detect regions of the genome under selection, we used three complementary approaches that together describe the evolutionary response of frog populations to Bd.
First, we used a multivariate linear mixed model to evaluate associations between population type (i.e., naive versus recovering) and individual variants.
For these single nucleotide polymorphisms (SNPs), we identified outlier variants using two Bonferroni-corrected significance levels: "liberal" = 0.05, "stringent" = 0.01.
Second, we used a splined window approach to evaluate larger genomic regions showing differentiation between population types in *F~ST~* and nucleotide diversity ($\pi$) <!--# Allie comment: Feels like we need to signal here that we are calculating the difference in pi between naive and persistent. pi-diff.  -->.
In this analysis, we identified outlier regions using a threshold for the W-statistic (4 standard deviations from the mean) and selected regions where outlier windows overlapped for *F~ST~* and $\pi$ <!--# Allie comment: Feels like we need to signal here that we are calculating the difference in pi between naive and persistent. pi - diff -->.
Third, we assessed whether putative genes or genomic regions that showed strong evidence of selection were associated with biological functions, molecular functions, or cellular processes (i.e., GO terms) consistent with possible resistance mechanisms.

@@

Our second analysis, using a splined window approach, identified regions of the genome that had both high differentiation (*F~ST~*) and a large difference in nucleotide diversity between naive and recovering populations ($\pi_{diff}$ = $\pi_{naive}$ - $\pi_{recovering}$).
We identified 58 outlier windows for *F~ST~* and 33 outlier windows for $\pi_{diff}$ (@fig-spline-manhattan A, B).
Of these, 9 regions were outliers for both metrics and two of these shared regions also contained outlier SNPs.<!--# "windows" and "regions" are used interchangeably in this paragraph. To avoid confusion, use "windows" throughout? --> A total of 33 annotated genes were found in the 9 shared regions (SI Dataset 2 <!--# In the supplementary_table_outliers_typofix.xlsx file > the splined_window_shared_ouliers sheet contains descriptions of 10 shared regions, not 9--> <!--# Allie comment: One is a repeat because there is a 2>1 match for one of the overlaps (i.e., One Fst outlier window overlaps with 2 pi outlier windows).  -->) and no GO terms were overrepresented within this set of genes.
The strongest signal for a reduction in $\pi$ in recovering populations (highest $\pi_{diff}$) <!--# Allie comment: For me this is enough of a reason to discuss this beyond the obvious relevance of the gene function.  -->, indicative of directional selection, occurred in a 163kb region on Contig19, 12.9Mb upstream of the RIN3 outlier SNP (@fig-spline-manhattan C).
This region contains \~500 SNPs and one annotated gene called "interferon-induced very large GTPase 1-like" (GVINP1).
Additionally, a shared outlier region on Contig1 contained two complement factor genes (C6 and C7).
Interestingly, this region had higher $\pi$ in the recovering populations, consistent with a pattern of balancing selection.
Finally, one shared outlier region on Contig8 contained one outlier SNP (TCF19) <!--# Allie comment: Again, this is a very good reason to discuss - there are both outlier windows and individuals SNPS here making it an important part outlier region to discuss BEYOND the cool MHC connection.  -->and was within 360kb of another outlier SNP (VARS) (@fig-spline-manhattan D).
This region (854kb from the beginning of the outlier windows to the VARS SNP) contained a total of 10 annotated genes.
In *Xenopus*, four of these genes occur in the extended MHC Class I region (FLOT1, TUBB, MDC1,TCF19) and three occur in the extended MHC Class II region (HSP70, LSM2, VARS) [@ohta2006].
Therefore, we conclude that this candidate region under selection is a part of the extended MHC Class I and II complex and shows synteny with other amphibian genomes.
<!--# Here or in the discussion we will need to place this result in the context of previous work on MHC in amphibians.  Readers not familiar with amphibian-Bd systems might miss the importance of this result --><!--# Based on Fu and Waldman (2017), are any of our data sufficient to conclude anything related to statement that "Bd-resistant species share similar pocket conformations within the MHC-II antigen-binding groove. Among susceptible species, survivors of epizootics bear alleles encoding these conformations."? What about MHC Class II characteristics linked to resistance by Bataille et al. (2015)? Savage and Zamudio (2011, 2016), Kosch et al. (2018)?-->

@@

@@

@@

Disease-induced declines are a central challenge for the conservation of global biodiversity, but strategies to recover affected species are generally lacking.
One potential strategy, and the one we use in the current study, leverages evolutionary rescue to recover species at the landscape scale.
Although previously proposed [@mendelson2019], to our knowledge this strategy has never been successfully implemented in wildlife disease systems [but see @joseph2018 for a small-scale example].
Specifically, we provide evidence that (i) MYL frogs have experienced strong selection following exposure to a recently-emerged lethal fungal disease (Bd), and (ii) individuals from these resistant populations can be used to reestablish viable populations despite ongoing Bd infection.
These results provide a rare, and perhaps novel, example of how evolutionary rescue can be used to successfully recover a species in the presence of disease.

We provide three lines of evidence that support evolutionary rescue as a primary mechanism of recovery in MYL frog systems.
First, our genomic analyses comparing naive and recovering MYL frog populations found notable differences in allele frequencies between adjacent Bd-naive and recovering <!--# Check on using "persistent" vs "recovering" vs "recovered" --> <!--# Roland response: Use "recovering" throughout -->populations, consistent with strong selection.
Importantly, some of the regions of the genome that showed signatures of selection are associated with known cellular and immunological mechanisms contributing to disease resistance in other species generally [@robledo2018; @robledo2020] and amphibians specifically [@rosenblum2008; @ellison2014] (for more information, see SI X).
For example, we identified a \~940kb region on contig 8 that shows evidence of selection in recovering populations and contains eight genes associated with either the MHC Class I or Class III regions<!--# Only Class I and II are mentioned in Results -->.
These results corroborate numerous previous studies linking MHC genes to Bd survival [e.g., @savage2011; @bataille2015].
Broadly, our genomic analyses imply that selection and rapid evolutionary change has contributed to survival in MYL frogs in the presence of Bd.
<!--# For brevity in the discussion, I tried to a pull a few key bullets out of Allie's text. Allie and Bree: Edit and expand as you see fit. -->

The second line of evidence that evolutionary rescue is a strong candidate mechanism for recovery is that repeated translocations and long-term monitoring of naturally recovered populations have demonstrated that resistant phenotypes are maintained beyond the natal environment in the presence of Bd (as reflected by the high year-one adult survival in 6 out of 12 translocated populations).
Moreover, resistance has a quantifiable phenotype in this system -- resistant, persisting populations have Bd infection intensities below the level known to cause significant host mortality [@vredenburg2010, @knapp2016].
Translocated populations maintain infection intensities below this threshold of infection intensity such that infection intensity is not a significant predictor of adult survival following translocations to new environments.
Finally, high adult survival in the presence of Bd persists into subsequent generations in non-natal environments, suggesting that resistance is heritable to some degree.
Although previous studies have documented signatures of selection in hosts following disease-induced declines [@epstein2016, @gignoux-wolfsohn2021], our study is unique in that these three lines of evidence provide a strong case that evolved resistance begets population recovery in both natal and non-natal habitats, consistent with evolutionary rescue.

@@

1\) STRENGTH OF SIGNAL: the gene regions we discuss specifically in our dataset are robust outliers.
In the case of the MHC region we have multiple SNPs in this region that are statistically significantly associated AND a larger pattern of depressed pi in the recovering pops and large Fst.
In the case of GVINP1 we have the only instance of two adjacent outlier spline windows for pi, and the largest difference in pi of any region we measured.
Finally for RIN3 the allele frequency maps tell the most compelling story and reflect the fact that this is the most statistically significant outlier SNP that has an annotation.
and

@@

### Selection Results
